Human Factors Influencing Visual Statistical Inference

### 2017-10-21 submitted-v1/sociology_paper.Rnw
cleaned the source

\documentclass[10pt]{article}
\usepackage{times}
%\usepackage{natbib}
%\usepackage{multicol}
\RequirePackage{natbib}
\usepackage[colorlinks=true, citecolor=blue, linkcolor=blue]{hyperref}
\usepackage{amsmath, amssymb, fullpage, amsthm, array, graphicx, asa, url}
%\usepackage[dvips]{graphics}

%\usepackage{hyperref} % for hyper reference

@@

% \usepackage{pifont} % this package is used to print check mark \checkmark
% \linespread{1.6} % factor 1.6 = double space

%\usepackage{setspace}
%\doublespacing

\usepackage{color}
\definecolor{mahbub}{rgb}{1.0, 0.6, 0.4}
%\usepackage[usenames,dvipsnames]{color}
\newcommand{\hh}[1]{{\color{magenta} #1}}
\newcommand{\dc}[1]{{\color{green} #1}}
\newcommand{\mm}[1]{{\color{mahbub} #1}}

@@

%\newtheorem{thm}{Theorem}[section]
%\newtheorem{dfn}{Definition}[section]
%\newtheorem{cor}{Corollary}[thm]
%\newtheorem{con}{Conjecture}[thm]
%\newtheorem{lemma}[thm]{Lemma}

%\topmargin -0.10in   % when making pdf
%\textheight 9.15in  % when making pdf

\pdfminorversion=4 % as instructed by JASA file upload

\begin{document}
%\tableofcontents

% Article top matter
\title{Human Factors Influencing Visual Statistical Inference }
\author{{Mahbubul Majumder, Heike Hofmann, Dianne Cook}
\thanks{Mahbubul Majumder is an Assistant Professor in the Department of Mathematics, University of Nebraska at Omaha, NE 68182 (e-mail: mmajumder@unomaha.edu), Heike Hofmann is Professor in the Department of Statistics and Statistical Laboratory, Iowa State University, Ames, IA 50011-1210, and  Dianne Cook is Professor in the Department of Econometrics and Business Statistics, Monash University, Australia. This research is supported in part by the National Science Foundation Grant \# DMS 1007697.}}
\date{\vspace{-.5in}}
%\date{\today}  %\today is replaced with the current date
\maketitle

\begin {abstract}  
Visual statistical inference is a way to determine significance of patterns found while  exploring data with graphics.  It relies on human observers inspecting a lineup: one  data plot randomly placed among a set of decoy plots. Each individual is different in their cognitive psychology and judiciousness, which can affect the visual inference. The usual way of estimating the effectiveness of a statistical test is its power. The estimate of power of a lineup is controlled by combining evaluations from multiple observers. Factors that may  affect the power of visual inference are the observers' demographics, visual skills, and experience, the sample of null plots taken from the null distribution, the position of the data plot in the lineup, and the signal strength in the data. This paper examines these factors. Results from multiple visual inference studies using Amazon's Mechanical Turk are examined to provide an assessment of these. The experiments suggest that individual skills vary substantially,  but demographics do not have a huge effect on performance. There is evidence that a learning effect exists but only in the sense that observers become faster with repeated evaluations, while accuracy stays the same. The placement of data plot in the lineup does not affect inferential results.

@@

%\begin{multicols}{2}
%\twocolumn

\section{Introduction}

@@

In visual inference, the test statistic is a plot of the observed data, called the {\it (actual) data plot}. To create a lineup the data plot is placed randomly in a grid of null plots. Null plots  are rendered from data generated as  specified by a null hypothesis. Often, this null hypothesis assumes independence or no structure in the data. An observer is then asked to evaluate the lineup. If the  actual data plot is detected by the observer, this counts as evidence against the null hypothesis. Based on the number of evaluations by independent observers and the number of times that the data plot is identified, we are  able to evaluate the significance of the graphical relationship in the same way and the same rigor as a conventional hypothesis test. % is rejected. This means that the structure in the actual data plot has significant structure, a pattern that is not simply due to randomness. Combining the choices of multiple observers provides more stability in the estimation of significance.

%\subsection{Visual inference example}

\begin{figure}[hbtp] 
   \centering
   \includegraphics[width=0.99\textwidth]{electoral-5-13.pdf} 
%      \includegraphics[width=0.99\textwidth]{electoral-1-1.pdf} 
   \caption{ \label{fig:elect} Which one of the plots, labelled 1 through 20 from top left to bottom right, is the most different from the others?}
\end{figure}

Figure~\ref{fig:elect} displays an example lineup, one of the plots is based on the observed data, while the remaining 19 plots show  data generated from a null model. Which one of the 20 plots is the most different from the others? 
When asked this question, 12 out of 72 observers picked the  data plot located in panel number $4^2-3$ in the lineup\footnote{The little piece of calculus used for describing the location of the data plot in the lineup provides a small cognitive hurdle that is supposed to enable the interested reader to inspect the lineup without being biased by knowing the location of the data plot.}.  The corresponding $p$-value is 0.0077, indicating sufficient evidence to reject the null hypothesis. 
% , with reasons given being 'asymmetry' (36\%), 'trend' (26\%) or `outliers' (16\%).

How do we interpret this finding, though? For that, we need to know the context of the data and we need to have more information about how the null plots were generated. 
This particular example investigates the results from the 2012 US presidential election in comparison to  poll results just prior to the election. 
Although this example is more simplistic than most of the tests conducted to date, it serves the purpose of illustrating the lineup protocol. The data consists of the difference in poll results between the two (major) presidential candidates, Obama and Romney, for all states. Each panel in Figure~\ref{fig:elect} shows an `electoral building' \citep{mosley2010} where each state in the union is represented by a rectangle. The difference in poll results is plotted horizontally, while the height of each box corresponds to the number of that state's electoral votes. Color indicates party affiliation.  The null hypothesis is that ``the election results are consistent with the polls''. Therefore the polling results provide the null model from which data is simulated. A normal model with mean and standard deviation based on a poll and its margin of error is used to simulate different scenarios that might have resulted on election day, if the polls were on target. Each null data set is generated as a set of draws from this model. % normal distribution, with mean equal to the difference in poll percentage of the latest state poll results, and standard deviation equal to 2.5, approximating a margin of error of 5\%. 
These samples are plotted as electoral buildings, and the plot based on the actual election results is placed randomly among the null plots in a lineup of size $5\times 4$. If the null hypothesis is true, the  data plot should look  like any of the other plots, and not be identifiable by an observer. Figure~\ref{fig:tower} shows the plot of the electoral building with the additional context information and labels.
 %A party needs 270 electoral votes or more to win the presidential election. 
%\begin{figure}[htbp] 
%   \centering
%   \includegraphics[width=0.8\linewidth]{electoral-5-13.pdf} 
%   \caption{Which plot looks the most different from the other plots?  Lineup \#1 - electoral building; another set of four lineups can be found in figs.~{fig:elect.2} through~{fig:elect.5}}
%   \label{fig:elect.1}
%\end{figure}

@@

%The rest of the 19 plots are generated from the assumption that there is no difference between two groups of data. For this, the group structure of the data is broken by random permutation of the groups.

%If the null hypothesis is true, the observed data should look similar to the rest of the plots in the lineup making it difficult to detect the observed plot. On the other hand, if the null hypothesis is not true, the observed data should be different from the rest of the plots in the lineup and observers should be able to pick the data plot out from the rest.

A lineup can be evaluated by a single person or multiple observers. A distribution similar to the binomial distribution, but adjusted for dependencies introduced by the lineup scenario, is used to calculate the $p$-value based on the number of times observers identify the actual data plot, which provides the information needed to make a decision on rejecting or failing to reject the null hypothesis \citep{vinference}.  To avoid expectation errors \citep{meilgaard}, particularly in emotionally charged areas such as election results, observers should not be aware of the data that constitutes a lineup, and should therefore not have seen the actual data plot before inspecting the lineup. This is the reason that we revealed the context of the election problem only after the interested reader had a chance to evaluate the lineup.%the election example, above, the scenario was explained after the lineup question, in the text.

When presenting a lineup to an observer  the accompanying question should be phrased in as general a manner as possible, 
%The question that is asked of the observer should be as general as possible, 
effectively asking the observer to pick the plot that is most different, and allowing observers to provide their own reasons for their choice. This ensures all possible deviations from the null hypothesis to be detected. 
However, if lineup tests are run in head-to-head comparisons with conventional statistical tests, such as the experiments  in  \cite{majumder:2013} or \cite{tengfei:2013}, these questions have to be phrased much more specifically, in order to avoid  type III errors \citep{mosteller:48} consisting of correctly rejecting the null hypothesis for a wrong reason. 
%In some of the studies, the ones described in  \cite{majumder:2013} very specific questions were asked, because the experiments were being conducted to compare results from the lineup protocol with those of conventional tests. 
In those experiments, the structure in the data is strictly controlled in the simulation process, which allows for specific questions to be asked. 
In the election example, observers were asked ``which plot is the most different?''. The type of plot, showing two (modified) stacked bar charts in different colors should suggest to the observer that the interesting feature is the difference between the two heights. 
Most observers based their decision on this -- besides  `asymmetry' between the towers (a reason suggested by us), most free-form responses for the reason of choice describe a comparison between the red and blue towers. 
In the responses for this lineup we see something that is very typical of responses in lineups: it is only a few panels from the lineup that observers pick, while most panels do not get picked or get picked only a few times. Out of a total of 72~responses, 23 observers picked panel~\#8, 12 observers picked the (data) panel \#13, 10 observers picked panel~\#4, 6 observers picked panel \#7, and all of the remaining panels were picked at most three times. From the panels picked, we see that the top three choices  are all extreme with respect to heights of blue versus red towers:  both panels \#8 and \#13 show a large difference between the height of the towers, while panel \#4 is the only panel, in which the red tower is higher than the blue one. 
% results from the turk study:
% panel:   1 10 11 12 13 14 15 16 18 19  2 20  3   4  5  6  7  8  9 
% #picks:  2  1    0  1 12    3  3    1  1   1  2   1  0 10  2  2  6 23  1

%Most observers got this, but it is possible that some observers might pick plot~4, where the red tower is slightly above the blue as the most different because it is the only plot with this feature \hh{(10 out of 72 observers actually did this)}. 
  %a better question may

% \hh{If we consider this difference to be the main objective of the visualization, a more focused question, that aligns more with a hypothesis test on the difference  in electoral votes  between parties, would} have been ``Which plot shows the biggest height difference between the two towers?'' %except that this tailors the inference to a specific feature which does not match the null hypothesis of interest.

% HH already included:
% For any evaluation the observer may or may not identify the actual data plot. Under the null hypothesis, the actual data plot should look similar to the null plots making it harder to detect. It is not expected that an observer would be able to detect the data plot in this scenario. But since there are limited number of plots in a lineup, which is 20 in the election example, there is a 1/20 chance that the observer would pick the actual plot.

The lineup protocols allows us to calculate all relevant properties that we are familiar with from conventional statistical tests. In particular, the power of a lineup is calculated as the {\it detection rate} at which observers identify the actual data plot. Visual power depends unlike power in conventional tests not only on the strength of the signal, but also on individuals' ability. 
%This proportion is associated with the Type~I error of the test. On the other hand, if null hypothesis is not true, the observed plot should look different from the null plots, making it easier to be detected. This is the definition of the power of the test. When multiple observers evaluate a lineup, the detection rate can be used to estimate the power. 
The ability of individual observers varies, and the effects that might influence this ability are the focus of this paper.

%\dc{*** Am trying to use detection rate rather than correct response. This is harder for this paper because we are evaluating subject's abilities to some extent, but we should try to use this in our vis inference papers.} % \mm{Di, do you suggest that I change ``correct response''  to ``detection rate'' in all the texts here?}

%For a lineup with fixed difficulty, the proportion correct responses may vary for different observers. If an observer evaluates multiple lineups, the proportion of correct evaluations by the observer could be used as a measure of performance by an observer. The another way of measuring the performance of the observer could be the time taken by an observer to evaluate a lineup. Since performance would also depend on the lineup difficulty it is essential to adjust for lineup difficulty before estimating the performance of the observer.

%Some of the factors that may affect the performance of the observer are discussed in Section~\ref{sec:factor_performance}. In this paper we focus on estimating the learning effect of the observer in terms of both proportion correct and time taken to evaluate a lineup through the attempts made during the evaluation of multiple plots. We also estimate the location effect of the observed data plot in a lineup. The other factors are controlled in simulation experiments so that learning effect and location effect can be estimated.

@@

%\subsection{Experiments pooled}

\section{Factors Potentially Affecting Visual Inference} \label{sec:factor_performance}

@@

%ased on human evaluation on a lineup a decision is made in visual statistical inference. While the visual test statistic is defined to have a direct influence on the observer to pick a plot based on the signal in the data, there are other human factors that may affect the observer performance. It is important to study the effect of those factors and examine the extent of their influence. This section presents a brief description of the factors that may affect the power of visual inference.

%\dc{You need to have more than one sentence to open this section, it needs a paragraph overviewing the factors.}

@@

%On the other hand, if the null hypothesis is not true, visual test statistic should not display any distinguishable pattern. In fact, some of the null plots may appear to be the most different plot in a lineup influencing the observer to chose a plot different than actual data plot. This is an elegant feature of the lineup. It force the observer to chose a wrong plot when the null hypothesis should not be rejected.

%Plots to do: for the same signal strength show the proportion correct for scatterplot and boxplot. eg., qplot(effect, ump-visual power, color= plot-type)

%\subsection{Question Design} 
\item{\bf Question Design:} A primary purpose of visual inference is to preclude the necessity to pre-specify discoverable features in data that is required by classical statistical inference. Visual inference supports discovery of structure, enabling the unexpected to surprise us.  This is tightly connected to the task that the human observer is asked to perform. For most purposes the observer is asked a very general question, such as to pick the plot that is most different from the others, and explain the reasons why they see it as different. By being general, all possible discoverable structures are included in the significance calculations \citep[see also ][]{buja:2009}.

@@

% The researcher knows about the underlying hypothesis but the observer does not necessarily know the underlying details of the lineup. So, the researcher needs to ask a question to the observer to answer while evaluating the lineup. This question should provide the observer a little clue so that the answer reflects the hypothesized patterns in the actual plot. For example Table~\ref{tbl:visual_stat} shows the questions asked for the simulation experiments considered for this study. Notice that for case 1 if the observer can identify the actual plot in the lineup that should indicate that the plot chosen has the most vertical difference between groups A and B which is exactly what the researchers intend to examine. Similarly for case 2, a correct evaluation would indicate that the slope is different than the slope that may show up just from randomness.

%These questions are very crucial for the power of visual test. They help observer think in a meaningful direction. Notice that there may be unnecessary patterns in the actual data plot which may not necessarily indicate the existence of the significant signal in the plot. These questions help observer not to be misguided by those patterns. To follow up further on this \citet{majumder:2013} also collected reasons of choice made by the observers and noticed that Type-III error may occur where the hypothesis may be correctly rejected but for a wrong reason.

%\subsection{Observer Personality}

%\subsection{Visual Perception of Human Eye} Different people observe the lineup in different ways.  With the help of eye-tracking equipment \citet{zhao:2012} track the observers' eyes to see how they go through the plots in a lineup to come to their answers. The result suggest that people have particular methods of reading lineups. Some people read lineups from left to write direction while some read from upward to downward. Some people start looking from the center of the lineup while others start from the top left corner. In the earlier phase of the exercise, the observer tend to scan the plots and then start comparing plots to make a final decision. Beside right-left or up-down directions observer show some diagonal movement too. 
%
%Given this pattern of human eye tracks, it may effect the performance of an observer depending on where the actual plot is placed in the lineup. If the actual plot is on the top left corner and the observer starts from that point it may be easier to detect the actual plot earlier in the exercise and the observer could get plenty of time to make comparison. Thus the position of the actual plot in a lineup may have some impact.

@@

%Some of the demographics of the observer such as age, gender, education level and geographical location may have effect on how an observer would examine the lineup. This may produce some variability in the performance of the observer. For example, a high school student may not respond same as a well educated person while evaluating a lineup. There may be variations in different age groups as well. Investigation on subjects with a well variety of age groups is necessary to study this variability.

%To meet the Institutional Review Board (IRB) requirement, it is necessary to exclude any subject less than 18 years of old to participate human subject experiment. Thus we intend to only focus on the performance of observer with age 18 years or older. We don't expect many older people to participate our study and we are interested of age from 18 to 65 years.

%Education level has some relation with age. Younger observer may not have higher degree. Thus some of the variability in performance for different education level may be confounded with age. On the other hand, it could be possible to have undergrad degree with higher age. Specially for any geographical location where not many certain gender group have higher education, this may occur. Thus the effect of all these demographical factors may be confounded at certain level.

%We don't expect much difference in performances between male and female observers. Gender may have some influence through the education level and geographical locations since some location may have small number of educated female population. We intend to include similar number of male and female subjects in our study so that this can be examined.

%\subsection{Learning Trend} 
\item{\bf Learning Trend:} One might expect that as an observer evaluates more lineups they  become more skillful in their evaluations. Each participant in  experiments~5, 6 and~7 was asked to evaluate a block of  ten lineups of the same type of data plot. The ten lineups were randomly chosen from the lineups  produced for each study. Before evaluating the first lineup, the observer needs to read instructions and become accustomed to the type of plot used in the lineup. In subsequent lineups  the type of plot is the same. It is possible that the observer becomes more skillful at recognizing the most different plot, either by more often detecting the actual data plot or more quickly reporting their choice. These two ways of measuring learning trend are evaluated on data collected from experiments~5, 6 and~7.

%When an observer evaluates a lineup he or she may learn something from the experiences of the earlier observations. If the same observer is shown another lineup the learning from previous evaluation may help. The more evaluations an observer makes the more skillful the observer may become. This learning trend may or may not be significant overall but it may influence the performance of the observer in some way.

%Learning may occur in two different ways. An observer can become more trained on the lineup structure and the pattern in the plots shown. This can be learning in proportion of correct evaluation. Or, he or she may become efficient in responding faster in the later evaluations. This can be observed as the time taken for evaluating a lineup. This paper investigates both of these learning trends with multiple simulation experiments.

%When an observer evaluate a lineup for the first time, he or she has to read through many instructions and become familiar with the environment of the experiment. We expect that this will require much more time compared to the rest of the evaluations. Time taken for the second or successive attempt should be much lower than the time taken for the first lineup. We attribute this to the adaptation to the experiment environment, not a learning of evaluating a lineup.

@@

%While evaluating the lineup some people may start looking from some specific part of the lineup.  In the earlier phase of the exercise, the observer tend to scan the plots and start comparing plots to make a final decision. Beside right-left or up-down directions observers show some diagonal movement too.

%Given a specific pattern of eye movement of the observer in examining the lineup, the location of actual plot in the lineup may have some effect. Those who start exploring from the left top corner may get first glance of the actual plot if it is on that location. This should give the observer more time to compare it with the rest of the null plots. Those who start looking from the center may scan towards right or left direction and thus don't have the chance to scan the null plots continuously as they could if they would start from a corner. Thus they should scan the center plot over and over again while scanning the left or right side plots. If the data plot is in the center location it may have multiple chance to be examined and hence be identified.

%It may be possible that some part of the lineup is less explored or even never scanned by the observer if he or she feels that the actual plot is found before even seriously scanning the whole lineup. In those scenarios the location where the observer first start scanning is important. Placing the actual data plot in that location may yield different results.

%\subsection{Sample of Null Plots} 
\item{\bf Sample of Null Plots:} In classical inference, the test statistic is compared to the full null distribution, to decide if it is extreme or not. In visual inference the actual plot is compared to plots of a finite number of samples from the null distribution. In the lineups used in the Turk experiments, the actual data plot is compared on 19 null plots. These null samples are random draws, and there is a chance that one or more null plots in the lineup may be similar or even more extreme than the actual data plot. The sample of null plots can affect the observer's decision. This is tested with the data from experiment~9, where the experiment was set up with lineups made from different null plots.

%each lineup, there is a sample of draws from a null distribution. A lineup becomes difficult to evaluate if one of the null plots appears to be very similar to the actual plot. When null hypothesis is true it is a common scenario. With alternative hypothesis being true,  it may happen if  we compare actual plot with many null plots. But we have only specific number of null plots in a lineup. So, a different sample of null plots may yield some variation in the difficulty of the lineup with the same actual plot. This may affect the performance of the observer while evaluating a lineup.

@@

\section{Experimental Methods}\label{sec:exp_design}

%While evaluating a lineup, the performance of the observer may depend on the factors described in Section~\ref{sec:factor_performance}. 
Two of the factors, signal in the data and individual abilities, were studied in \cite{majumder:2013}. The choice of visual test statistic was examined in \cite{heike:2012}. In each of these analyses demographic factors were given a cursory glance, to ensure that they did not have large effects on the results. The design of experiments 5, 6 and 7 enables the examination of learning trend, which is studied in this paper. Experiment~9 was a real test case for visual statistical inference, and in order to understand the significance of the structure in the genomic data, multiple lineups were made in which location of the actual data plot, and the sample of nulls, were randomized. This enables the assessment of the effect of these factors on the results. This section describes the experimental methods used to examine the effects of demography, placement of the actual data plot, sample of null plots and the existence of a learning trend.

% but for the most part data collected to examine demographic effects . In this paper we intend to study effect of the factors such as human demographics, learning trend, actual plot locations and selection of null plots. This section presents the simulation experiment setup, data collection methods and data analysis plans with models to asses the influence of those factors on visual statistical inference.

%\dc{Learning trend and location effect need to be separate sections, cannot be simply bolded.}

\subsection{Demographic Factors}  %In addition to obtain evaluation responses of the lineups, all the 10 experiments shown in Table~\ref{tbl:visual_stat} were designed to collect the following demographic information of the subjects participating the experiments:
\noindent
For all  ten experiments shown in Table~\ref{tbl:visual_stat}, the following demographic information was collected from subjects:

>>>>

\documentclass[12pt]{article}

\usepackage{times}
\RequirePackage{natbib}
\usepackage[colorlinks=true, citecolor=blue, linkcolor=blue]{hyperref}
\usepackage{amsmath, amssymb, fullpage, amsthm, array, graphicx, asa, url}

@@

% JASA guideline to have double spaced fonts
\usepackage{setspace}
\doublespacing

\usepackage{color}
\definecolor{mahbub}{rgb}{1.0, 0.6, 0.4}

\newcommand{\hh}[1]{{\color{magenta} #1}}
\newcommand{\dc}[1]{{\color{green} #1}}
\newcommand{\mm}[1]{{\color{mahbub} #1}}

@@

\pdfminorversion=4 % as instructed by JASA file upload
% NOTE: To produce blinded version, replace "1" with "0" below. 
% Blinded version will also remove acknowledgement at the bottom.
\newcommand{\blindDoc}{1}

\begin{document}

% ===================================================================================
%Article top matter 
%------------------------------------------------------------------------------------

\if1\blindDoc
{
  \title{Human Factors Influencing Visual Statistical Inference}
  \author{{Mahbubul Majumder, Heike Hofmann, Dianne Cook}
\thanks{Mahbubul Majumder is an Assistant Professor in the Department of Mathematics, University of Nebraska at Omaha, NE 68182 (e-mail: mmajumder@unomaha.edu), Heike Hofmann is Professor in the Department of Statistics and Statistical Laboratory, Iowa State University, Ames, IA 50011-1210, and  Dianne Cook is Professor in the Department of Econometrics and Business Statistics, Monash University, Australia. This research is supported in part by the National Science Foundation Grant \# DMS 1007697.}}
\date{\vspace{-.5in}}
  \maketitle
} \fi

\if0\blindDoc
{
  \bigskip
  \bigskip
  \bigskip
  \begin{center}
    {\bf \LARGE Human Factors Influencing Visual Statistical Inference}
\end{center}
  \medskip
} \fi

% ===================================================================================
\begin {abstract} 
%------------------------------------------------------------------------------------

Visual statistical inference is a way to determine significance of patterns found while  exploring data with graphics.  It relies on human observers inspecting a lineup: one  data plot randomly placed among a set of decoy plots. Each individual is different in their cognitive psychology and judiciousness, which can affect the visual inference. The usual way of estimating the effectiveness of a statistical test is its power. The estimate of power of a lineup is controlled by combining evaluations from multiple observers. Factors that may  affect the power of visual inference are the observers' demographics, visual skills, and experience, the sample of null plots taken from the null distribution, the position of the data plot in the lineup, and the signal strength in the data. This paper examines these factors. Results from multiple visual inference studies using Amazon's Mechanical Turk are examined to provide an assessment of these. The experiments suggest that individual skills vary substantially,  but demographics do not have a huge effect on performance. There is evidence that a learning effect exists but only in the sense that observers become faster with repeated evaluations, while accuracy stays the same. The placement of data plot in the lineup does not affect inferential results.

@@

% ===================================================================================
\section{Introduction}  
%------------------------------------------------------------------------------------

@@

In visual inference, the test statistic is a plot of the observed data, called the {\it (actual) data plot}. To create a lineup the data plot is placed randomly in a grid of null plots. Null plots  are rendered from data generated as  specified by a null hypothesis. Often, this null hypothesis assumes independence or no structure in the data. An observer is then asked to evaluate the lineup. If the  actual data plot is detected by the observer, this counts as evidence against the null hypothesis. Based on the number of evaluations by independent observers and the number of times that the data plot is identified, we are  able to evaluate the significance of the graphical relationship in the same way and the same rigor as a conventional hypothesis test.

\begin{figure}[hbtp] 
   \centering
   \includegraphics[width=0.99\textwidth]{electoral-5-13.pdf} 
   \caption{ \label{fig:elect} Which one of the plots, labelled 1 through 20 from top left to bottom right, is the most different from the others?}
\end{figure}

Figure~\ref{fig:elect} displays an example lineup, one of the plots is based on the observed data, while the remaining 19 plots show  data generated from a null model. Which one of the 20 plots is the most different from the others? 
When asked this question, 12 out of 72 observers picked the  data plot located in panel number $4^2-3$ in the lineup\footnote{The little piece of calculus used for describing the location of the data plot in the lineup provides a small cognitive hurdle that is supposed to enable the interested reader to inspect the lineup without being biased by knowing the location of the data plot.}.  The corresponding $p$-value is 0.0077, indicating sufficient evidence to reject the null hypothesis.

How do we interpret this finding, though? For that, we need to know the context of the data and we need to have more information about how the null plots were generated. 
This particular example investigates the results from the 2012 US presidential election in comparison to  poll results just prior to the election. 
Although this example is more simplistic than most of the tests conducted to date, it serves the purpose of illustrating the lineup protocol. The data consists of the difference in poll results between the two (major) presidential candidates, Obama and Romney, for all states. Each panel in Figure~\ref{fig:elect} shows an `electoral building' \citep{mosley2010} where each state in the union is represented by a rectangle. The difference in poll results is plotted horizontally, while the height of each box corresponds to the number of that state's electoral votes. Color indicates party affiliation.  The null hypothesis is that ``the election results are consistent with the polls''. Therefore the polling results provide the null model from which data is simulated. A normal model with mean and standard deviation based on a poll and its margin of error is used to simulate different scenarios that might have resulted on election day, if the polls were on target. Each null data set is generated as a set of draws from this model. % normal distribution, with mean equal to the difference in poll percentage of the latest state poll results, and standard deviation equal to 2.5, approximating a margin of error of 5\%. 
These samples are plotted as electoral buildings, and the plot based on the actual election results is placed randomly among the null plots in a lineup of size $5\times 4$. If the null hypothesis is true, the  data plot should look  like any of the other plots, and not be identifiable by an observer. Figure~\ref{fig:tower} shows the plot of the electoral building with the additional context information and labels.

@@

A lineup can be evaluated by a single person or multiple observers. A distribution similar to the binomial distribution, but adjusted for dependencies introduced by the lineup scenario, is used to calculate the $p$-value based on the number of times observers identify the actual data plot, which provides the information needed to make a decision on rejecting or failing to reject the null hypothesis \citep{vinference}.  To avoid expectation errors \citep{meilgaard}, particularly in emotionally charged areas such as election results, observers should not be aware of the data that constitutes a lineup, and should therefore not have seen the actual data plot before inspecting the lineup. This is the reason that we revealed the context of the election problem only after the interested reader had a chance to evaluate the lineup.

When presenting a lineup to an observer the accompanying question should be phrased in as general a manner as possible, effectively asking the observer to pick the plot that is most different, and allowing observers to provide their own reasons for their choice. This ensures all possible deviations from the null hypothesis to be detected. However, if lineup tests are run in head-to-head comparisons with conventional statistical tests, such as the experiments in \cite{majumder:2013} or \cite{tengfei:2013}, these questions have to be phrased much more specifically, in order to avoid  type III errors \citep{mosteller:48} consisting of correctly rejecting the null hypothesis for a wrong reason. In those experiments, the structure in the data is strictly controlled in the simulation process, which allows for specific questions to be asked.

In the election example, observers were asked ``which plot is the most different?''. The type of plot, showing two (modified) stacked bar charts in different colors should suggest to the observer that the interesting feature is the difference between the two heights. Most observers based their decision on this -- besides  `asymmetry' between the towers (a reason suggested by us), most free-form responses for the reason of choice describe a comparison between the red and blue towers. In the responses for this lineup we see something that is very typical of responses in lineups: it is only a few panels from the lineup that observers pick, while most panels do not get picked or get picked only a few times. Out of a total of 72~responses, 23 observers picked panel~\#8, 12 observers picked the (data) panel \#13, 10 observers picked panel~\#4, 6 observers picked panel \#7, and all of the remaining panels were picked at most three times. From the panels picked, we see that the top three choices  are all extreme with respect to heights of blue versus red towers:  both panels \#8 and \#13 show a large difference between the height of the towers, while panel \#4 is the only panel, in which the red tower is higher than the blue one.

The lineup protocols allows us to calculate all relevant properties that we are familiar with from conventional statistical tests. In particular, the power of a lineup is calculated as the {\it detection rate} at which observers identify the actual data plot. Visual power depends unlike power in conventional tests not only on the strength of the signal, but also on individuals' ability. The ability of individual observers varies, and the effects that might influence this ability are the focus of this paper.

@@

% ===================================================================================
\section{Factors Potentially Affecting Visual Inference} \label{sec:factor_performance} 
%------------------------------------------------------------------------------------

@@

@@

\item{\bf Question Design:} A primary purpose of visual inference is to preclude the necessity to pre-specify discoverable features in data that is required by classical statistical inference. Visual inference supports discovery of structure, enabling the unexpected to surprise us.  This is tightly connected to the task that the human observer is asked to perform. For most purposes the observer is asked a very general question, such as to pick the plot that is most different from the others, and explain the reasons why they see it as different. By being general, all possible discoverable structures are included in the significance calculations \citep[see also ][]{buja:2009}.

@@

@@

\item{\bf Learning Trend:} One might expect that as an observer evaluates more lineups they  become more skillful in their evaluations. Each participant in  experiments~5, 6 and~7 was asked to evaluate a block of  ten lineups of the same type of data plot. The ten lineups were randomly chosen from the lineups  produced for each study. Before evaluating the first lineup, the observer needs to read instructions and become accustomed to the type of plot used in the lineup. In subsequent lineups  the type of plot is the same. It is possible that the observer becomes more skillful at recognizing the most different plot, either by more often detecting the actual data plot or more quickly reporting their choice. These two ways of measuring learning trend are evaluated on data collected from experiments~5, 6 and~7.

@@

\item{\bf Sample of Null Plots:} In classical inference, the test statistic is compared to the full null distribution, to decide if it is extreme or not. In visual inference the actual plot is compared to plots of a finite number of samples from the null distribution. In the lineups used in the Turk experiments, the actual data plot is compared on 19 null plots. These null samples are random draws, and there is a chance that one or more null plots in the lineup may be similar or even more extreme than the actual data plot. The sample of null plots can affect the observer's decision. This is tested with the data from experiment~9, where the experiment was set up with lineups made from different null plots.

@@

% ===================================================================================
\section{Experimental Methods}\label{sec:exp_design}
%------------------------------------------------------------------------------------

Two of the factors, signal in the data and individual abilities, were studied in \cite{majumder:2013}. The choice of visual test statistic was examined in \cite{heike:2012}. In each of these analyses demographic factors were given a cursory glance, to ensure that they did not have large effects on the results. The design of experiments 5, 6 and 7 enables the examination of learning trend, which is studied in this paper. Experiment~9 was a real test case for visual statistical inference, and in order to understand the significance of the structure in the genomic data, multiple lineups were made in which location of the actual data plot, and the sample of nulls, were randomized. This enables the assessment of the effect of these factors on the results. This section describes the experimental methods used to examine the effects of demography, placement of the actual data plot, sample of null plots and the existence of a learning trend.

\subsection{Demographic Factors}
\noindent
For all  ten experiments shown in Table~\ref{tbl:visual_stat}, the following demographic information was collected from subjects:

\begin{enumerate} \leftmargin 5cm  \itemsep 0in
\item {\it Age group}, with categories set to be 18-25, 26-30, 31-35, 36-40, 41-45, 46-50, above 50.
\item {\it Gender}, male or female.
\item {\it Education level}, with levels being high school or less, some undergraduate courses, undergraduate degree, some graduate courses, and graduate degree.
\item {\it Geographical location}, collected from the IP address of the participants' computer, as latitude, longitude, city and country. 
\end{enumerate}

Let $Y_{ij}$ denote the response from observer $i$ on a lineup $j$, with $Y_{ij} =1$ if the actual data plot is chosen, otherwise $Y_{ij} =0$. The factors are examined in association with the observer's response using a logistic regression with random effect terms:

\begin{equation} \label{eqn:demographic_response}
g( \pi_{ij} )= \mu + \alpha_{k(i)} + \gamma_{l(i)} + \tau_{m(i)}+ \kappa_{s(i)} + \ell_j,  
\end{equation}

where $\pi_{ij}=  E(Y_{ij})$ is the probability that  observer $i$ picks the actual data plot from lineup $j$, $\mu$ is an overall population average, $\alpha$, $\gamma$, $\tau$ and  $\kappa$ are the effects of age group $k(i)$, gender $l(i)$, education level $m(i)$ and country name $s(i)$, respectively, for observer $i$. The term $\ell_j$ is a random intercept predicting lineup difficulty level and we assume independence and normality of the errors, i.e.\ $\ell_j \sim N(0, \sigma_\ell^2)$.  $g(.)$ denotes the {\it logit} link function $g(\pi)=\log(\pi) - \log(1-\pi); 0 \le \pi \le 1$.
